---
title: "Corncob_higher_taxlevels"
author: "J. Meyer"
date: "2024-12-10"
output: html_document
---

```{r setup, include=FALSE}
library(phyloseq)
library(corncob)
library(plyr)
library(dplyr)
writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
```

### Prepare data for corncob analysis: BOTH CORAL SPECIES, FULL TENT ONLY


```{r, echo=FALSE}
# read in count data
otu <- read.table("Tent_Expm_silva_nochloronomito_otu_table.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("Tent_Expm_silva_nochloronomito_taxa_table.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("Tent_Expm_Metadata.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps  # 11845 taxa and 96 samples

ps1 = subset_samples(ps, Tent.treatment == "FT")
ps1 # 11845 taxa and 36 samples, keeping both coral species here
ps1 <- prune_taxa(taxa_sums(ps1) > 1, ps1) #remove taxa not found in remaining samples
ps1 # 4936 taxa and 36 samples, keeping both coral species here
# Now export cleaned otu and taxa tables from phyloseq for future reference
otu = as(otu_table(ps1), "matrix")
taxon = as(tax_table(ps1), "matrix")
metadata = as(sample_data(ps1), "matrix")
write.table(otu,"silva_nochloronomito_otu_table_ps1.txt",sep="\t",col.names=NA)
write.table(taxon,"silva_nochloronomito_taxa_table_ps1.txt",sep="\t",col.names=NA)
write.table(metadata,"metadata_ps1.txt",sep="\t",col.names=NA)

##### In both the otu table and the taxa table, replace the ASV sequence with an ASV number

otu <- read.table("silva_nochloronomito_otu_table_ps1_corncob.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps1_corncob.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps1.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps1 <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps1 #4936 taxa and 36 samples


# collapse ASVs into higher taxonomic groups for DA analysis
ps1_family <- ps1 %>% phyloseq::tax_glom("Family")
ps1_family # 549 taxa (families) and 36 samples

ps1_order <- ps1 %>% phyloseq::tax_glom("Order")
ps1_order # 336 taxa (orders) and 36 samples

ps1_class <- ps1 %>% phyloseq::tax_glom("Class")
ps1_class # 147 taxa (classes) and 36 samples


```

### Now use corncob to look for differentially abundant taxa: ASV LEVEL, BOTH CORAL SPECIES, FULL TENT ONLY

# results: 17 differential abundant ASV out of 4936; 8 ASVs enriched After low oxygen treatment = 5 ASVs Clostridia, 1 ASV Bacteroidia, 1 ASV Gammaproteobacteria, 1 Campylobacteria


```{r, echo=FALSE}

ps1 # 4936 ASVs and 36 samples


# DA determined relative to first reference level, so set this manually to the "before" treatment
sample_data(ps1)$Expm.status<-factor(sample_data(ps1)$Expm.status,levels=c("Bef","Aft","Rec"))


set.seed(1)
timepoint.da <- differentialTest(formula = ~ Expm.status, 
                                 phi.formula = ~ Expm.status,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Expm.status,
                                 test = "Wald", boot = FALSE,
                                 data = ps1,
                                 fdr_cutoff = 0.05)

timepoint.da$significant_taxa # 17 differential abundant ASV out of 4936

#quick look at results with corncob's plotting feature
plot(timepoint.da, level="Genus") # lists by genus

# create a function that will return the values of interest
# This one is for when there is only one comparison (here I only want to know what changed after low oxygen)
extractmods <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                       "Std.Error" = model$coefficients[2, 2], 
                       "p" = model$coefficients[2, 4])
  return(result)
}

# save the timepoint.da output and add asv, p.adj values to it, and ultimately taxonomy info
timepoint.da.models <- lapply(timepoint.da$significant_models, extractmods)
names(timepoint.da.models) <- timepoint.da$significant_taxa

# Add ASVs to the taxonomy table and save the significant asvs
tax_table(ps1)[,7] <- rownames(tax_table(ps1))
sig.taxonomy.timepoint <- as.data.frame(tax_table(ps1)[timepoint.da$significant_taxa,]) 

# Move the data from a list to a dataframe and add taxonomy info
timepoint.da.models.df <- ldply(timepoint.da.models, data.frame) %>% 
  left_join(sig.taxonomy.timepoint, by = c(".id" = "sequence")) %>%
  mutate(family_asv = paste0(Family, "_(", .id,")"))

### Save these data so you don't have to re-run the model ###
write.table(timepoint.da.models.df, "DA_ASV_timepoint.txt", sep="\t",row.names = FALSE)

```

### differentially abundant taxa: FAMILY LEVEL, BOTH CORAL SPECIES, FULL TENT ONLY

# results: 29 differential abundant families out of 549; 13 families enriched after low oxygen

```{r, echo=FALSE}

ps1_family #549 families and 36 samples


# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(ps1_family)$Expm.status<-factor(sample_data(ps1_family)$Expm.status,levels=c("Bef","Aft","Rec"))


set.seed(1)
timepoint.da <- differentialTest(formula = ~ Expm.status, 
                                 phi.formula = ~ Expm.status,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Expm.status,
                                 test = "Wald", boot = FALSE,
                                 data = ps1_family,
                                 fdr_cutoff = 0.05)

timepoint.da$significant_taxa # 29 differential abundant families out of 549

#quick look at results with corncob's plotting feature
plot(timepoint.da) # lists full taxonomy
plot(timepoint.da, level="Family") # lists just Family

# create a function that will return the values of interest
# This one is for when there is only one comparison
extractmods <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                       "Std.Error" = model$coefficients[2, 2], 
                       "p" = model$coefficients[2, 4])
  return(result)
}

# save the timepoint.da output and add asv, p.adj values to it, and ultimately taxonomny info
timepoint.da.models <- lapply(timepoint.da$significant_models, extractmods)
names(timepoint.da.models) <- timepoint.da$significant_taxa

# Add ASVs to the taxonomy table and save the significant asvs
tax_table(ps1_family)[,7] <- rownames(tax_table(ps1_family))
sig.taxonomy.timepoint <- as.data.frame(tax_table(ps1_family)[timepoint.da$significant_taxa,]) 

# Move the data from a list to a dataframe and add taxonomy info
timepoint.da.models.df <- ldply(timepoint.da.models, data.frame) %>% 
  left_join(sig.taxonomy.timepoint, by = c(".id" = "sequence"))

### Save these data so you don't have to re-run the model ###
write.table(timepoint.da.models.df, "DA_FAMILY_timepoint.txt", sep="\t",row.names = FALSE)
```



### differentially abundant taxa: ORDER LEVEL, BOTH CORAL SPECIES, FULL TENT ONLY

# results: 32 differential abundant orders out of 336; 12 orders enriched after low oxygen

```{r, echo=FALSE}

ps1_order # 336 orders and 36 samples

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(ps1_order)$Expm.status<-factor(sample_data(ps1_order)$Expm.status,levels=c("Bef","Aft","Rec"))


set.seed(1)
timepoint.da <- differentialTest(formula = ~ Expm.status, 
                                 phi.formula = ~ Expm.status,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Expm.status,
                                 test = "Wald", boot = FALSE,
                                 data = ps1_order,
                                 fdr_cutoff = 0.05)

timepoint.da$significant_taxa # 32 differential abundant orders out of 336

#quick look at results with corncob's plotting feature
plot(timepoint.da) # lists full taxonomy
plot(timepoint.da, level="Order") # lists just Order

# create a function that will return the values of interest
# This one is for when there is only one comparison
extractmods <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                       "Std.Error" = model$coefficients[2, 2], 
                       "p" = model$coefficients[2, 4])
  return(result)
}

# save the timepoint.da output and add asv, p.adj values to it, and ultimately taxonomny info
timepoint.da.models <- lapply(timepoint.da$significant_models, extractmods)
names(timepoint.da.models) <- timepoint.da$significant_taxa

# Add ASVs to the taxonomy table and save the significant asvs
tax_table(ps1_order)[,7] <- rownames(tax_table(ps1_order))
sig.taxonomy.timepoint <- as.data.frame(tax_table(ps1_order)[timepoint.da$significant_taxa,]) 

# Move the data from a list to a dataframe and add taxonomy info
timepoint.da.models.df <- ldply(timepoint.da.models, data.frame) %>% 
  left_join(sig.taxonomy.timepoint, by = c(".id" = "sequence"))

### Save these data so you don't have to re-run the model ###
write.table(timepoint.da.models.df, "DA_ORDER_timepoint.txt", sep="\t",row.names = FALSE)

```




### differentially abundant taxa: CLASS LEVEL, BOTH CORAL SPECIES, FULL TENT ONLY

# results: 20 differential abundant classes out of 147; 8 classes enriched after low oxygen


```{r, echo=FALSE}

ps1_class # 147 classes and 36 samples

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(ps1_class)$Expm.status<-factor(sample_data(ps1_class)$Expm.status,levels=c("Bef","Aft","Rec"))


set.seed(1)
timepoint.da <- differentialTest(formula = ~ Expm.status, 
                                 phi.formula = ~ Expm.status,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Expm.status,
                                 test = "Wald", boot = FALSE,
                                 data = ps1_class,
                                 fdr_cutoff = 0.05)

timepoint.da$significant_taxa # 20 differential abundant classes out of 147

#quick look at results with corncob's plotting feature
plot(timepoint.da) # lists full taxonomy
plot(timepoint.da, level="Class") # lists just Class


# create a function that will return the values of interest
# This one is for when there is only one comparison (here I only want to know what changed after low oxygen)
extractmods <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                       "Std.Error" = model$coefficients[2, 2], 
                       "p" = model$coefficients[2, 4])
  return(result)
}

# save the timepoint.da output and add asv, p.adj values to it, and ultimately taxonomny info
timepoint.da.models <- lapply(timepoint.da$significant_models, extractmods)
names(timepoint.da.models) <- timepoint.da$significant_taxa

# Add ASVs to the taxonomy table and save the significant asvs
tax_table(ps1_class)[,7] <- rownames(tax_table(ps1_class))
sig.taxonomy.timepoint <- as.data.frame(tax_table(ps1_class)[timepoint.da$significant_taxa,]) 

# Move the data from a list to a dataframe and add taxonomy info
timepoint.da.models.df <- ldply(timepoint.da.models, data.frame) %>% 
  left_join(sig.taxonomy.timepoint, by = c(".id" = "sequence"))

### Save these data so you don't have to re-run the model ###
write.table(timepoint.da.models.df, "DA_CLASS_timepoint.txt", sep="\t",row.names = FALSE)

```



### Corncob DA analysis with coral species separated


```{r, echo=FALSE}
#prepare data by coral species 
psS = subset_samples(ps1, Species == "Aten")
psS # 4936 taxa and 18 samples, just ATEN
psS <- prune_taxa(taxa_sums(psS) > 1, psS) #remove taxa not found in remaining samples
psS # 2024 taxa and 18 samples, just ATEN
otu = as(otu_table(psS), "matrix")
taxon = as(tax_table(psS), "matrix")
metadata = as(sample_data(psS), "matrix")
write.table(otu,"silva_nochloronomito_otu_table_psS_ATEN_FT.txt",sep="\t",col.names=NA)
write.table(taxon,"silva_nochloronomito_taxa_table_psS_ATEN_FT.txt",sep="\t",col.names=NA)
psS = subset_samples(ps1, Species == "Ssid")
psS # 4936 taxa and 18 samples, just SSID
psS <- prune_taxa(taxa_sums(psS) > 1, psS) #remove taxa not found in remaining samples
psS # 3613 taxa and 18 samples, just SSID
otu = as(otu_table(psS), "matrix")
taxon = as(tax_table(psS), "matrix")
metadata = as(sample_data(psS), "matrix")
write.table(otu,"silva_nochloronomito_otu_table_psS_SSID_FT.txt",sep="\t",col.names=NA)
write.table(taxon,"silva_nochloronomito_taxa_table_psS_SSID_FT.txt",sep="\t",col.names=NA)




#### Corncob with only ATEN
psA # 2024 ASVs and 18 samples

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(psA)$Expm.status<-factor(sample_data(psA)$Expm.status,levels=c("Bef","Aft","Rec"))


set.seed(1)
timepoint.da <- differentialTest(formula = ~ Expm.status, 
                                 phi.formula = ~ Expm.status,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Expm.status,
                                 test = "Wald", boot = FALSE,
                                 data = psA,
                                 fdr_cutoff = 0.05)

timepoint.da$significant_taxa # 11 differential abundant ASVs out of 2024

#quick look at results with corncob's plotting feature
plot(timepoint.da) # lists full taxonomy
plot(timepoint.da, level="Genus") # lists just genus

# create a function that will return the values of interest
# This one is for when there is only one comparison (here I only want to know what changed after low oxygen)
extractmods <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                       "Std.Error" = model$coefficients[2, 2], 
                       "p" = model$coefficients[2, 4])
  return(result)
}

# save the timepoint.da output and add asv, p.adj values to it, and ultimately taxonomny info
timepoint.da.models <- lapply(timepoint.da$significant_models, extractmods)
names(timepoint.da.models) <- timepoint.da$significant_taxa

# Add ASVs to the taxonomy table and save the significant asvs
tax_table(psA)[,7] <- rownames(tax_table(psA))
sig.taxonomy.timepoint <- as.data.frame(tax_table(psA)[timepoint.da$significant_taxa,]) 

# Move the data from a list to a dataframe and add taxonomy info
timepoint.da.models.df <- ldply(timepoint.da.models, data.frame) %>% 
  left_join(sig.taxonomy.timepoint, by = c(".id" = "sequence"))

### Save these data so you don't have to re-run the model ###
write.table(timepoint.da.models.df, "DA_ASV_timepoint_ATEN.txt", sep="\t",row.names = FALSE)

########## ATEN 11 DA ASVS, 4 ASVs enriched after low oxygen = Fusibacter ASVs 6,9,17,18




#### Corncob with only SSID
psS # 3613 ASVs and 18 samples

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(psS)$Expm.status<-factor(sample_data(psS)$Expm.status,levels=c("Bef","Aft","Rec"))


set.seed(1)
timepoint.da <- differentialTest(formula = ~ Expm.status, 
                                 phi.formula = ~ Expm.status,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Expm.status,
                                 test = "Wald", boot = FALSE,
                                 data = psS,
                                 fdr_cutoff = 0.05)

timepoint.da$significant_taxa # 5 differential abundant ASVs out of 3613, Fusibacter ASV18 and Peptostr. ASV20 enriched after low oxygen

#quick look at results with corncob's plotting feature
plot(timepoint.da) # lists full taxonomy
plot(timepoint.da, level="Genus") # lists just genus

# create a function that will return the values of interest
# This one is for when there is only one comparison (here I only want to know what changed after low oxygen)
extractmods <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                       "Std.Error" = model$coefficients[2, 2], 
                       "p" = model$coefficients[2, 4])
  return(result)
}

# save the timepoint.da output and add asv, p.adj values to it, and ultimately taxonomny info
timepoint.da.models <- lapply(timepoint.da$significant_models, extractmods)
names(timepoint.da.models) <- timepoint.da$significant_taxa

# Add ASVs to the taxonomy table and save the significant asvs
tax_table(psS)[,7] <- rownames(tax_table(psS))
sig.taxonomy.timepoint <- as.data.frame(tax_table(psS)[timepoint.da$significant_taxa,]) 

# Move the data from a list to a dataframe and add taxonomy info
timepoint.da.models.df <- ldply(timepoint.da.models, data.frame) %>% 
  left_join(sig.taxonomy.timepoint, by = c(".id" = "sequence"))

### Save these data so you don't have to re-run the model ###
write.table(timepoint.da.models.df, "DA_ASV_timepoint_SSID.txt", sep="\t",row.names = FALSE)



```
